Exploring Anomalies in Pacific Northwest Special Education Trends

### 2020-12-07 scripts/Final_project.Rmd
Numerize numbers
-Sooke

@@

```{r rural play, echo = FALSE}

##Looking into Sooke (SD 62 - pop. 9,269) categorized under Victoria so urban but technically rural
## In progress

sooke <- district_class %>%
  filter(DISTRICT_NAME == "Sooke") %>%
  as_tibble()

ggplot(sooke, aes(SPECIAL_NEEDS_CODE)) +
  facet_wrap(~SCHOOL_YEAR)
  geom_col()
```

>>>>

 # Making student number column numeric
x <- c(10)
BCdata[ , x] <- apply(BCdata[ , x], 2,
                               function(x)as.numeric(as.character(x)))

@@

### 2020-12-07 scripts/Final_project.tex
-Sooke

Educational policy can impact practice by reorganizing educational delivery and reconceptualizing the meaning and impact of education on society (Ball, 1998). Educational policy exists as a mechanism to drive changes in school buildings that result in socially important outcomes (e.g., increasing literacy acquisition through improved instruction, reducing instances of bullying through prevention programs). However, the relationship between policy and practice in Western formal systems of education is often unclear, misdirected, or atheoretical (Cohen et al., 2009). Leveraging the power of policy as a change agent for the delivery and conceptualization of education stands to benefit a significant proportion of students (Swanson, 2007).

When policy is enacted, the intended outcomes are often not realized for months, years, or decades (Gu et al., 2018). For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores (Luke \& McArdle, 2009). Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest (Michelmore \& Dynarski, 2017).

@@

We intend to explore several questions regarding longitudinal trends where educational policy may have impacted special education practice. First, we will use descriptive statistics to describe trends in disability prevalence over time. The initial analyses will inform a discussion on developmental trajectories by illustrating shifts in prevalence growth from static to linearly increasing based on age of diagnosis for the Oregon data, which can serve as a springboard to make inferences about BC data (e.g., common age of onset for diagnosis per category and diagnostic patterns over time). In studying the differences and similarities between OR and BC datasets, we will also engage in a discussion on diagnostic terminology across regions with respect to the definitions as detailed by the Diagnostic and Statistical Manual of Mental Disorders (DSM; American Psychological Association, 2013), in part as a response to a challenge set by differing terminology between BC and OR categorization (i.e., the United States federal guidelines for exceptionality designation are similar but not identical to exceptionality designation guidelines in BC).

Second, we hope to closely analyze any changes, or lack thereof, within the BC data following Canadian Supreme Court rulings between the British Columbia Teachers Federation and the province of BC (See Lambert, 2017). Canadian Supreme Court rulings have significant implications on educational policy in the provinces and, in 2016, resulted in a limit to the number of students with disability designations in classrooms resulting in an expansion of the number of specialist teachers' that school districts are required to hire. The driving theory of change is that decreasing student-to-teacher ratios and capping the number of students with exceptionalities in each classroom causes student designation practices to change. Theoretically, school district administrators made decisions based on current staffing levels, student designations, diagnostic capacity and malleability, as well as external pressures to adjust practice according to legal mandates.

@@

@@

Taken side by side, a simple visualization detailing student exceptionality designation presents the overall similarities between urban and rural areas, as in figures x and x below. Of note, the \enquote{Gifted} designation suffers a severe drop around 2005/2006 for urban and rural areas alike, though the change is more pronounced within rural districts.

@@

\includegraphics{Final_project_files/figure-latex/rural play-1.pdf} \includegraphics{Final_project_files/figure-latex/rural play-2.pdf} \includegraphics{Final_project_files/figure-latex/rural play-3.pdf}

@@

The rapid decrease in students identified as gifted in rural districts between the years of 2004 and 2007 cannot be explained by one sole factor in this exploratory analysis. With respect to the speculative nature of this assertion, the authors assume that educational policy at the provincial level impacts educational practice across school districts. In 2006, the British Columbia Ministry of Education released Special education services: A manual of policies, procedures and guidelines, which explicitly states the acceleration programming that should be afforded to students with exceptional academic talent (Kanevsky \& Clelland, 2016). In fact, this 2006 version of the manual was a revision of the original 1995 manual, and this initial version included scant guidance on service provision to gifted students. This 2016 shift in policy links to a higher degree of resources afforded to gifted students through \enquote{independent, guided education} (p.~52). Given the additional resources needed to properly implement gifted educational services, rural school districts may have de-designated students and sought other ways to provide instruction that meets the needs of their students. For example, Lo et al.~(2019) suggest that the British Columbia Ministry of Education's recent curriculum update makes amends for the traditional test-and-place gifted student support model by giving educators the tools and inclusive framework to offer a rich, meaningful education to all students with exceptionalities.

@@

\hypertarget{refs}{}

>>>>

Educational policy can impact practice by reorganizing educational delivery and reconceptualizing the meaning and impact of education on society (Ball, 1998). Educational policy exists as a mechanism to drive changes in school buildings that result in socially important outcomes (e.g., increasing literacy acquisition through improved instruction, reducing instances of bullying through prevention programs). However, the relationship between policy and practice in Western formal systems of education is often unclear, misdirected, or atheoretical (Cohen, McCabe, Michelli, \& Pickeral, 2009). Leveraging the power of policy as a change agent for the delivery and conceptualization of education stands to benefit a significant proportion of students (Swanson, 2007).

When policy is enacted, the intended outcomes are often not realized for months, years, or decades ({\textbf{???}}). For example, changing school district policy on professional development to align with research-informed practices may be constructed to increase student academic achievement; though, there is insufficient evidence pointing to how long professionalized intervention with teachers will take to increase student academic scores (Luke \& McArdle, 2009). Longitudinal data helps to understand the impact of policy change because of the year-over-year comparisons that illustrate differences, inflection points, or anomalies of interest (Michelmore \& Dynarski, 2017).

@@

We intend to explore several questions regarding longitudinal trends where educational policy may have impacted special education practice. First, we will use descriptive statistics to describe trends in disability prevalence over time. The initial analyses will inform a discussion on developmental trajectories by illustrating shifts in prevalence growth from static to linearly increasing based on age of diagnosis for the Oregon data, which can serve as a springboard to make inferences about BC data (e.g., common age of onset for diagnosis per category and diagnostic patterns over time). In studying the differences and similarities between OR and BC datasets, we will also engage in a discussion on diagnostic terminology across regions with respect to the definitions as detailed by the Diagnostic and Statistical Manual of Mental Disorders (DSM, Association \& others, 2013), in part as a response to a challenge set by differing terminology between BC and OR categorization (i.e., the United States federal guidelines for exceptionality designation are similar but not identical to exceptionality designation guidelines in BC).

Second, we hope to closely analyze any changes, or lack thereof, within the BC data following Canadian Supreme Court rulings between the British Columbia Teachers Federation and the province of BC (see Lambert, 2017). Canadian Supreme Court rulings have significant implications on educational policy in the provinces and, in 2016, resulted in a limit to the number of students with disability designations in classrooms resulting in an expansion of the number of specialist teachers' that school districts are required to hire. The driving theory of change is that decreasing student-to-teacher ratios and capping the number of students with exceptionalities in each classroom causes student designation practices to change. Theoretically, school district administrators made decisions based on current staffing levels, student designations, diagnostic capacity and malleability, as well as external pressures to adjust practice according to legal mandates.

@@

\caption{\label{tab:Provincial Data Table}Special Needs Designation by Select Years (BC)}

\begin{tabular}{lllllll}
\toprule
Special Needs & \multicolumn{1}{c}{1996/1997} & \multicolumn{1}{c}{2001/2002} & \multicolumn{1}{c}{2005/2006} & \multicolumn{1}{c}{2010/2011} & \multicolumn{1}{c}{2015/2016} & \multicolumn{1}{c}{2019/2020}\\
\midrule
Physically Dependent & 2,080.00 & 2,233.00 & 1,953.00 & 1,725.00 & 1,521.00 & 1,636.00\\
Deafblind & 126.00 & 92.00 & 96.00 & 131.00 & 182.00 & 163.00\\
Mod To Profound Intellectual Disability & 6,522.00 & 7,512.00 & 7,572.00 & 6,450.00 & 5,382.00 & 4,732.00\\
Phys Disability Or Chronic Health Impair & 6,722.00 & 14,291.00 & 19,088.00 & 23,896.00 & 30,872.00 & 35,969.00\\
Visual Impairment & 1,203.00 & 1,136.00 & 1,057.00 & 860.00 & 844.00 & 792.00\\
Deaf Or Hard Of Hearing & 4,416.00 & 4,652.00 & 4,322.00 & 3,578.00 & 3,505.00 & 3,805.00\\
Autism Spectrum Disorder & 1,913.00 & 4,487.00 & 7,942.00 & 18,054.00 & 31,423.00 & 52,475.00\\
Inten Behav Inter/Serious Mental Illness & 19,855.00 & 23,419.00 & 26,970.00 & 26,526.00 & 27,470.00 & 33,156.00\\
Severe Learning Disabilities & NA & 55,194.00 & NA & NA & NA & NA\\
Mild Intellectual Disability & NA & 12,701.00 & 8,965.00 & 7,152.00 & 6,966.00 & 8,341.00\\
Behavior Disorder - Moderate & NA & 26,707.00 & NA & NA & NA & NA\\
Behavior Disorder - Rehabilitation & NA & 18,726.00 & NA & NA & NA & NA\\
Gifted & NA & 64,665.00 & 43,997.00 & 27,560.00 & 19,736.00 & 19,717.00\\
Learning Disability & NA & NA & 62,913.00 & 68,944.00 & 74,709.00 & 84,703.00\\
Moderate Behaviour Support/Mental Illnes & NA & NA & 34,311.00 & 21,949.00 & 20,396.00 & 24,867.00\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} NAs indicate masked data where students <10}
\end{tablenotes}

@@

Taken side by side, a simple visualization detailing student exceptionality designation presents the overall similarities between urban and rural areas, as in figures 7 and 8 below. Of note, the \enquote{Gifted} designation suffers a severe drop around 2005/2006 for urban and rural areas alike, though the change is more pronounced within rural districts.

@@

\includegraphics{Final_project_files/figure-latex/rural play-1.pdf}

@@

The rapid decrease in students identified as gifted in rural districts between the years of 2004 and 2007 cannot be explained by one sole factor in this exploratory analysis. With respect to the speculative nature of this assertion, the authors assume that educational policy at the provincial level impacts educational practice across school districts. In 2006, the British Columbia Ministry of Education released Special education services: A manual of policies, procedures and guidelines, which explicitly states the acceleration programming that should be afforded to students with exceptional academic talent (Kanevsky \& Clelland, 2013). In fact, this 2006 version of the manual was a revision of the original 1995 manual, and this initial version included scant guidance on service provision to gifted students. This 2016 shift in policy links to a higher degree of resources afforded to gifted students through \enquote{independent, guided education} (p.~52). Given the additional resources needed to properly implement gifted educational services, rural school districts may have de-designated students and sought other ways to provide instruction that meets the needs of their students. For example, Lo et al. (2019) suggest that the British Columbia Ministry of Education's recent curriculum update makes amends for the traditional test-and-place gifted student support model by giving educators the tools and inclusive framework to offer a rich, meaningful education to all students with exceptionalities.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-american_2013}{}%
Association, A. P., \& others. (2013). \emph{Diagnostic and statistical manual of mental disorders (dsm-5)}. American Psychiatric Pub.

\leavevmode\hypertarget{ref-ball_1998}{}%
Ball, S. J. (1998). Big policies/small world: An introduction to international perspectives in education policy. \emph{Comparative Education}, \emph{34}(2), 119--130.

\leavevmode\hypertarget{ref-cohen_2009}{}%
Cohen, J., McCabe, L., Michelli, N. M., \& Pickeral, T. (2009). School climate: Research, policy, practice, and teacher education. \emph{Teachers College Record}, \emph{111}(1), 180--213.

\leavevmode\hypertarget{ref-kanevsky_2013}{}%
Kanevsky, L. S., \& Clelland, D. (2013). Accelerating gifted students in canada: Policies and possibilities. \emph{Canadian Journal of Education}, \emph{36}(3), 229--271.

\leavevmode\hypertarget{ref-lambert_2017}{}%
Lambert, S. (2017). Public education: Reflections on fifteen years of public policy. \emph{BC Studies: The British Columbian Quarterly}, (194), 163--173.

\leavevmode\hypertarget{ref-lo_2019}{}%
Lo, C. O., Tan, Y. S. M., Chrostowski, M., Hu, S.-F., Chan, D., Sue, D. M., \ldots{} Li, W. (2019). Reenvisioning gifted education in british columbia: A qualitative research protocol of policy analysis in the context of a redesigned curriculum. \emph{International Journal of Qualitative Methods}, \emph{18}, 1609406918822233.

\leavevmode\hypertarget{ref-luke_2009}{}%
Luke, A., \& McArdle, F. (2009). A model for research-based state professional development policy. \emph{Asia-Pacific Journal of Teacher Education}, \emph{37}(3), 231--251.

\leavevmode\hypertarget{ref-michelmore_2017}{}%
Michelmore, K., \& Dynarski, S. (2017). The gap within the gap: Using longitudinal data to understand income differences in educational outcomes. \emph{AERA Open}, \emph{3}(1), 2332858417692958.

\leavevmode\hypertarget{ref-swanson_2007}{}%
Swanson, J. D. (2007). Policy and practice: A case study of gifted education policy implementation. \emph{Journal for the Education of the Gifted}, \emph{31}(2), 131--164.

\leavevmode\hypertarget{ref-walker_2013}{}%
Walker, J., \& Bergmann, H. von. (2013). Teacher education policy in canada: Beyond professionalization and deregulation. \emph{Canadian Journal of Education}, \emph{36}(4), 65--92.
